### 2018-05-28 write-up/implementation.tex
Fix minor layout issues
Fix many small writing errors

\prechapter{%
    I will describe the structure of LT4LA: a DSL (domain specific language)
    \textbf{library} written in OCaml that transpiles to OCaml. I will explain
    the features of its core language and show how readable (equational and
    algebraic expressions) and easy to reason about linear-algebra programs
    can be elaborated into memory-efficient ones in the core language that are
    then checked for safety (with respect to aliasing, read/write permissions,
    memory allocation, re-use and de-allocation). Finally, I will explain how
    such programs can be transpiled to OCaml code that is not obviously safe.
    Although I refer to OCaml-specific features of the type-checker and
    transpiler, I believe the ideas described in this chapter can easily be
    implemented in other languages and are also general enough to transpile to
    different back-end languages, such as C or Fortran.
}%

@@

LT4LA follows the structure of a typical compiler for a (E)DSL\@. From the
start, I made a concerted effort to (1) write pure-functional code (typically
using a monadic-style) which helped immensely with modularity and debugging
when tests showed errors (2) produce readable, useful and precise
error-messages in the hope that someone who did not understand linear types
could still use the LT4LA (3) write tests and set-up
continuous-integration for all non-trivial functions so that I could spot and
correct errors that were not caught by OCaml's type system whenever I
implemented new features or refactored my code.

@@

    \item \textbf{Parsing \& Desugaring}. A generated, LR(1) parser parses a text
        file into a syntax tree, which is then desugared into a smaller, more
        concise abstract syntax tree. The former aims to mimic OCaml syntax
        with a few extensions and keywords so that it is familiar and thus easy
        to pick-up for OCaml users. The latter allows for the type-checker to
        be simpler to implement and easier to specify. In general, this part
        will vary for different languages or can be dealt with differently
        using combinators (the EDSL approach) or a syntax-extension if the host
        language offers such support.

    \item \textbf{Type-checking}. The abstract syntax tree is explicitly-typed,
        with some inference to make it less verbose and more convenient to
        write typical programs.

    \item \textbf{Matrix Expressions}. During type-checking, if a matrix-expression
        is encountered, it is either successfully elaborated into an expression
        in the abstract syntax tree which is then consequently type-checked, or
        fails to find suitable routines to calculate the given expression.

    \item \textbf{Code Generation}. The abstract syntax tree is translated into
        standard OCaml and a few-particular `optimisations' are made to produce
        more readable code. This process is type-preserving: the linear type
        system is embedded into OCaml's type system, and so when the OCaml compiler
        compiles the generated code, it acts as a sanity check on the code produced.

@@

A full description of the core language can be found in
Appendix~\ref{chap:ott_spec}. For convenience, I have reproduced its type
grammar and defined how I interpret those types in OCaml's type system in
Figure~\ref{fig:type_grammar}.  The point of LT4LA is to keep the same values
and types we are familiar with, but \emph{restrict their usage} in certain
ways. LT4LA's core language's main features are: intuitionistic values,
value-restriction, fractional-capabilities (inferrable at call sites),
if-expressions and recursion.

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{Type grammar of LT4LA (left) and an informal interpretation of it
        into OCaml types (right).}\label{fig:type_grammar}
\end{figure}

@@

To make LT4LA a usable DSL, I needed some way of using values zero or more than
once, as a programmer would an intuitionistic value. For this, I added the
!-constructor at the type-level of the DSL and the \ltfla{Many}-constructor and
the \ltfla{let Many <id> = .. in .. } eliminator at the term-level of the DSL.
The idea behind the !-type is that the value uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a !-type. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to
emulate intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to
\ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>}
(similarly for function argument bindings). The reader can check (using the
rules in Appendix~\ref{chap:ott_spec}) that this has the effect of moving
\ltfla{x : !t} from the linear to the intuitionistic environments, only if
\ltfla{<exp> : !t}.

However, just that desugaring alone is not enough to prevent a user from taking
an array or matrix and moving it into the intuitionistic environments. Why?
There are certain situations in which we \emph{should not} use the \ltfla{Many}
constructor.  Consider the following code: \ltfla{let Many x = Many (array 5)
in <body>}; the expression \ltfla{array 5} uses no linearly-typed variables
from the linear environment. Although we could just reject types of the form
\ltfla{!(_ arr)} to fix this simple example, what about pairs \ltfla{let Many
xy = Many (3, array 5) in <body>}?  Ad-hoc pattern matching on the type cannot
account for all possible situations. With the last case, we can use \ltfla{xy}
as many times as we would like, destruct the pair to get the second component
and thus create \emph{distinct} read-write aliases to the same array.  Alas,
now arrays can be used intuitionistically and all the benefits of linearity are
lost.  Or are they?

@@

Not quite, but to understand how we can fix this problem, we need to question
an assumption left implicit up until this point: what does \ltfla{Many} even
mean in the DSL? How is it translated to OCaml? What is the OCaml runtime
behaviour of the DSL's \ltfla{Many}-constructor? One option is to go down the
C++ route and make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a
runtime reference-count for arrays. I chose to not go for this option because
it went against the \emph{explicitness} and \emph{predictability} that C and
Fortran have. It would make analysing when and what is allocated and freed more
like the higher-level languages I was trying to move away from.

My aim is to show linear types are simple to understand and apply to
linear-algebra libraries, enough so that it can be grafted (in a limited way)
on to existing languages as a \emph{library}. In that spirit, the simplest
thing that the DSL's \ltfla{Many}-constructor can mean in the OCaml runtime is
\emph{nothing}.  This language construct is translated into a standard OCaml
type-and-constructor of the form \ltfla{type 'a bang = Many of 'a [@@unboxed]}.
The \emph{unboxed} annotation means that the type and its constructor only
exist for the purpose of type-checking in OCaml; \emph{the OCaml runtime
representation of values of type \ltfla{'a} and \ltfla{'a bang} is exactly the
same}.

With this understanding, our problem is that arrays and matrices are unlike
other values such as integers and elements because (under the OCaml hood)
calling a function with an array argument copies a \emph{pointer} to the array
rather than the array itself, instead of the \emph{value} itself. So, we can
start making a distinction, \emph{defining} elements, integers, booleans,
intuitionistic variables, units and lambda-expressions that capture no linear
variables as \emph{values} (since they cannot break referential-transparency)
and anything else (arrays, matrices, expressions which can be reduced, such as
function application and if-expressions) as not being `values'. If this sounds
familiar, it is because this is the same \emph{value-restriction} `trick' from
the world of polymorphic types applied to linearity instead. We then have the
rule that in LT4LA, we can only use \ltfla{Many} on expressions that are
defined to be \emph{values} and \emph{use no linear variables}.

@@

I started off with linearity and showing how it helps a programmer keep track
of of memory allocation and deallocation, and then demonstrated how to add
intuitionism for the only values we wish to be intuitionistic. I will now
explain how I enforce safe aliasing and read/write permissions.

Array and matrix types are parameterised by
\emph{fractional-capabilities}~\cite{boyland}.  A fraction of 1 ($2^0$)
represents complete ownership of a value; in particular, this allows a
programmer to write or free it. Creating an array gives you ownership of it;
the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents `0').
Once you have ownership of an array, you can free it: \ltfla{free : z arr --o
unit}.  Importantly, because a linear-value may only be used once, the array
just freed is \emph{out of scope} for following expressions, preventing
use-after-free.  Ownership also enables you to write to the array: \ltfla{set :
z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just sugar
for \ltfla{set w i j}). Here, linearity prevents accessing aliases which
represented the array \emph{before} the mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
means a program can \emph{appear} to use a variable multiple times, important
for keeping LT4LA usable and readable. The underscore is how a programmer tells
the compiler to automatically \emph{infer} the correct fractional-capability
based on the other arguments passed to the function. In conjunction with the
requirement that function declarations need type-annotations for their
arguments, \textbf{this allows a fractional-capability to be \emph{correctly
inferred} in \emph{any} LT4LA expression}.

Fractions exist to provide both safe aliasing and read/write permissions, via
the primitives \ltfla{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former, two
arrays returned (which happen to just be the given array) can now only be read
from and not written to. If you want to write to this array, you must use the
latter to combine other read-only aliases until you are left with a value of
type \ltfla{z arr}, guaranteeing no other aliases exists.

Given this set-up, a programmer now \emph{statically} has \emph{perfect}
information about aliasing and ownership of values in the program. They can
only write to or free an array only when they own it; ownership guarantees no
other aliases exist. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
information can be used to write more readable (equational and algebraic) code
using easy to reason about, value-semantic expressions whose \emph{intensional}
behaviour (regarding aliasing, read/write permissions and memory allocation,
re-use and de-allocation)  is also explicitly and precisely described and
\emph{automatically checkable}. Now the programmer need not resort to manually
figuring out and inserting \texttt{noalias} annotations and worrying about what
variables can and cannot be written to, re-used or freed; instead they can let
the loyal and tireless compiler do the heavy lifting.

@@

An if-expression's condition may evaluate either way at run time, so a
programmer must guarantee that both branches use the same set of linear
variables. Writing the type-checker in a pure-functional monadic style paid off
here because I could now sandwich monadic values with state-adjustments either
side of it. Given two monadic values that represented type-checking two
branches of an if-expression, I could use the code in
Figure~\ref{fig:same_resources} to easily save, reset and compare the state
either side of running those monadic values.

@@

    \caption{Implementation of same\_resources helper method for type-checking
        if-expressions. Note the monadic style helped compose computations that
        affected the type-checker's state in a simple manner.}\label{fig:same_resources}

@@

A non-recursive function may be used more than once if it does not refer to any
linear variables from the surrounding scope. So, we can desugar something like
\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
let Many f =  Many (fun y : !int -> x + y) in  <body>}.  Recursion is slightly
more complicated: we can desugar the factorial function
(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int, if
(*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many}, must also
not use any linear variables from its surrounding scope, because calling a
recursive function may evaluate its body multiple times.

@@

In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
implementing a \emph{Kalman filter}~\cite{kalman}, a powerful set of equations
(Figure~\ref{fig:maths_kalman}) applicable to a wide variety of problems. From
the comments, we see that every variable is annotated with the step/matrix
expression that it will hold at some point during the computation (an
equivalent alternative, say in C++, could be to have a meaningful name for each
step and manually keep track of which names alias the same location).

@@

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed or written-over;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking values are neither used after they are freed/written-over \emph{nor} leaked.
\end{itemize}

Indeed, an inexperienced programmer could take the  na\"ive approach of just
copying sub-expressions by default and then letting the compiler tell it which
copies are never used and removing them systematically until it type-checks.
While it is not quite a black-box, push-button compilation of an expression, I
would argue that, it is just as easy (if not easier) to become familiar with as
Rust and its borrow-checker.

\begin{figure}[tp]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA. In contrast with the Fortran
        (Figure~\ref{fig:fortran_kalman}) or C (Figure~\ref{fig:cblas_kalman})
        implementations, this one is readable (equational and algebraic),
        using easy to reason about, value-semantic expressions. In addition to
        that, its \emph{intensional} behaviour(regarding aliasing, read/write
        permissions and memory allocation, re-use and de-allocation) is also
        explicitly and precisely described and (unlike with Fortran or C)
        \emph{automatically checkable}.}\label{fig:ltfla_kalman}
\end{figure}

@@

All of the syntax in Figure~\ref{fig:ltfla_kalman} can be unambiguously
desugared \emph{before} type-checking, through fairly simple pattern matching.
Matrix expression compilation is well-trodden territory in
academia~\cite{rocklin_thesis, fabregat_thesis, gunnels_flame, linnea, taco}
but this is, to my knowledge, the \textbf{first type-based approach} to it.

@@

One of the first hurdles I encountered was compositionality: to compile $AB +
C$ it is typically better to use a BLAS routine directly rather than but first
compiling $A$, then $B$, then adding a call to multiply them, then compiling
$C$ and finishing with a call to add the results.

Another compositionality problem is that a call to a linear function does not
just return a result, but a sequence of re-bindings that dictate which
variables are still in scope/usable.  As such, to compile an expression, you
need to provide a CPS-style function of type $var \rightarrow exp$ representing
how you would use a variable representing the result in the rest of the
expression.

@@

To make the code produced readable, I added a few `optimisations', to the
compiler, to `re-sugar' some of the constructs translated where appropriate.
Almost all of them involved the erasing the !-eliminator which is not needed in
regular, intuitionistic OCaml. So, so we can simply replace any expression of
the form:

@@

    \item \ltfla{let xy, z = <exp> in let x,y = xy in <body>} with
        \ltfla{let (x,y), z = <exp in <body>},

@@

\begin{figure}[tp]
    \centering
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let rec f i n x0 row =
  if Prim.extract @@ Prim.eqI i n then (row, x0)
  else
    let row, x1 = Prim.get row i in
    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
in
f
    \end{minted}
    \caption{Recursive OCaml function for a summing an array, generated (at
        \emph{compile time}) from the code in Figure~\ref{fig:ltfla_sumarray},
        passed through \texttt{ocamlformat} for presentation.}\label{fig:ocaml_sumarray}

@@

It is clear that both Figures~\ref{fig:fortran_kalman}
and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the whole
behaviour and intent (with respect to read/writes permissions, memory
management and aliasing) of the program and the BLAS primitives it uses more
accurately than a Fortran, C or OCaml implementation could.

@@

Advantages of generating code at compile time include avoiding runtime overhead
as well as catching any interface changes (type-signatures now include
\emph{intensional} information about which parameters are read from or written
to) between the generated code and the code that uses it via OCaml's type
system, \emph{another benefit} (in addition to a sanity check on LT4LA's
type-checker implementation) of \emph{embedding} LT4LA's type system inside of
OCaml's.

I suspect that this approach will be very valuable to not just \emph{users} of
libraries such as Numpy or Owl, but also their \emph{implementors}: library
functions which present a safe, value-semantic interface but use unsafe,
mutating operations on the inside could now be expressed using LT4LA and have
an extra safety-net and automatic checking for their implementations.

\section{Summary}

>>>>

\prechapter{%
    I will describe the structure of LT4LA: a DSL (domain specific language)
    \textbf{library} written in OCaml that transpiles to OCaml. I will explain
    the features of its core language and show how readable (equational and
    algebraic expressions) and easy to reason about linear algebra programs
    can be elaborated into memory-efficient ones in the core language that are
    then checked for safety (with respect to aliasing, read/write permissions,
    memory allocation, re-use and de-allocation). Finally, I will explain how
    such programs can be transpiled to OCaml code that is not obviously safe.
    Although I refer to OCaml-specific features of the type checker and
    transpiler, I believe the ideas described in this chapter can easily be
    implemented in other languages and are also general enough to transpile to
    other back-end languages, such as C or Fortran.
}%

@@

LT4LA follows the structure of a typical compiler for a (E)DSL\@. From the
start, I made a concerted effort to (1) write pure-functional code (typically
using a monadic-style) which helped immensely with modularity and debugging
when tests showed errors (2) produce readable, useful and precise
error-messages in the hope that someone who did not understand linear types
could still use LT4LA (3) write tests and set-up continuous-integration for all
non-trivial functions so that I could spot and correct errors that were not
caught by OCaml's type system whenever I implemented new features or refactored
my code.

@@

    \item \textbf{Parsing}. A generated, LR(1) parser parses a text file into a
        syntax tree. I tried to mimic OCaml syntax modulo a few extensions and
        keywords so that it is familiar and thus easy to pick-up for OCaml
        users. In general, this part will vary for different languages and can
        also be dealt with using combinators or syntax-extensions (the EDSL
        approach) if the host language offers such support .

    \item \textbf{Desugaring}. The syntax tree is then desugared into a
        smaller, more concise, abstract syntax tree. This allows for the type
        checker to be simpler to specify and easier to implement.

    \item \textbf{Matrix Expressions} are also desugared into the abstract
        syntax tree through simple, yet effective pattern-matching.

    \item \textbf{Type checking}. The abstract syntax tree is explicitly typed,
        with some inference to make writing typical programs more convenient.

    \item \textbf{Code Generation}. The abstract syntax tree is translated into
        OCaml, with a few `optimisations' to produce more readable code. This
        process is type-preserving: LT4LA's type system is embedded into
        OCaml's (Figure~\ref{fig:type_grammar}), and so the OCaml type checker
        acts as a sanity check on the generated code.

@@

A full description of the core language can be found in
Appendix~\ref{chap:ott_spec}. For convenience, I have reproduced its type
grammar and its translation into OCaml in Figure~\ref{fig:type_grammar}.  The
point of LT4LA is to keep the same values and types we are familiar with from
OCaml, but \emph{restrict their usage}. LT4LA's core language's main features
are: intuitionistic values, value-restriction, fractional-capabilities
(inferrable at call sites), if-expressions and recursion.

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{Type grammar of LT4LA (left) and my translation of it
        into OCaml (right).}\label{fig:type_grammar}
\end{figure}

@@

To make LT4LA a usable DSL, I needed a way of allowing values to either never
be used or be used more than once. For this, I added the \mbox{!-constructor}
at the type-level of the DSL and its corresponding introduction (the
\ltfla{Many}-constructor) and elimination (the \ltfla{let Many <id> = .. in ..
} destructor) forms at the term-level of the DSL.  The idea behind the
\mbox{!-constructor} is that a value of that type uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a \mbox{!-type}. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to emulate
intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to \ltfla{let Many x =
<exp> in let Many x = Many (Many x) in <body>} (similarly for function argument
bindings). The reader can check (using the rules in
Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.

However, just that desugaring alone is not enough to prevent a user from taking
an array or matrix and moving it into the intuitionistic environments. Why?
There are certain situations in which we \emph{should not} use the \ltfla{Many}
constructor.  Consider the following code: \ltfla{let Many x = Many (array 5)
in <body>}; the expression \ltfla{array 5} uses no linearly-typed variables
from the environment. Although we could just reject types of the form
\ltfla{!(_ arr)} to fix this simple example, what about pairs \ltfla{let Many
xy = Many (3, array 5) in <body>}?  Ad-hoc pattern matching on the type cannot
account for all possible situations. With the last case, we can use \ltfla{xy}
as many times as we would like, destruct the pair to get the second component
and thus create \emph{distinct} read-write aliases to the same array.  Alas,
now arrays can be used intuitionistically and all the benefits of linearity are
lost.  Or are they?

@@

Not quite, but to understand how we can fix this problem, we need to question
an assumption left implicit up until this point: what does \ltfla{Many} even
mean in the DSL? How is it translated to OCaml? What is the OCaml runtime
behaviour of the DSL's \ltfla{Many}-constructor? One option is to go down the
C++ route and make \ltfla{Many} behave like a \ltfla{shared_ptr} and act as a
runtime reference-count for arrays. I chose to not go for this option because
it went against the \emph{explicitness} and \emph{predictability} that C and
Fortran have. It would make analysing when and what is allocated and freed more
like the higher-level languages I was trying to move away from.

My aim is to show linear types are simple to understand and apply, enough so
that they can be grafted (in a limited way) on to existing languages as a
\emph{library}. In that spirit, the simplest thing that the DSL's
\ltfla{Many}-constructor can mean in the OCaml runtime is \emph{nothing}.
LT4LA's !/\ltfla{Many} constructors are translated into OCaml
\ltfla{bang}/\ltfla{Many} constructors declared as \ltfla{type 'a bang = Many
of 'a [@@unboxed]}.  The \emph{unboxed} annotation means that the type and its
constructor only exist for the purpose of type checking in OCaml; \emph{the
OCaml runtime representation of values of type \ltfla{'a} and \ltfla{'a bang}
is exactly the same}.

With this understanding, our problem is that arrays and matrices are unlike
other values such as integers and elements because (under the OCaml hood)
calling a function with an array argument copies a \emph{pointer} to the array
rather than the array itself. So, we can start making a distinction,
\emph{defining} elements, integers, booleans, intuitionistic variables, units
and lambda-expressions that capture no linear variables as \emph{values} (since
they cannot break referential-transparency) and anything else (arrays,
matrices, expressions which can be reduced, such as function application and
if-expressions) as not being `values'. If this sounds familiar, it is because
this is the same \emph{value-restriction} `trick' from the world of polymorphic
types applied to linearity instead. We then have the rule that in LT4LA, we can
only use \ltfla{Many} on expressions that are defined to be \emph{values} and
\emph{use no linear variables}.

@@

I started off with linearity and showed how it helps a programmer keep track
of memory allocation and deallocation, and then demonstrated how to add
intuitionism for the only values we wish to be intuitionistic. I will now
explain how I enforce safe aliasing and read/write permissions.

Array and matrix types are parameterised by
\emph{fractional-capabilities}~\cite{boyland}.  A fraction of 1 ($2^0$)
represents complete ownership of a value. Creating an array gives you ownership
of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents
`0').  Once you have ownership of an array, you can free it: \ltfla{free : z
arr --o unit}.  Importantly, because a linear value may only be used once, the
array just freed is \emph{out of scope} for following expressions,
\textbf{preventing use-after-free}.  Ownership also enables you to write to the
array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i]
:= j} is just sugar for \ltfla{set w i j}). Here, linearity prevents accessing
aliases which represented the array \emph{before} the mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
means a program can \emph{appear} to use a variable multiple times, important
for keeping LT4LA usable and readable. The underscore is how a programmer tells
the compiler to automatically \emph{infer} the correct fractional-capability
based on the other arguments passed to the function. In conjunction with the
requirement that function declarations need type-annotations for their
arguments, \textbf{this allows fractional-capabilities to be \emph{correctly
inferred} in \emph{any} LT4LA expression}.

Fractions exist to provide both safe aliasing and read/write permissions, via
the primitives \ltfla{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former,
the two arrays returned (which are just aliases of the given array) can now
only be read from and not written to. If you want to write to this array, you
must use the latter to combine other read-only aliases until you are left with
only one value of type \ltfla{z arr}, guaranteeing no other aliases exists.

Given this set-up, a programmer now \emph{statically} has \emph{perfect}
information about aliasing and ownership of values in the program. They can
only write to or free an array only when they own it; ownership guarantees no
other aliases exist. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
information can be used to write more readable (equational and algebraic) code
using easy to reason about, value-semantic expressions whose \emph{intensional}
behaviour and assumptions (regarding aliasing, read/write permissions and
memory allocation, re-use and de-allocation) are also explicitly and precisely
described and \emph{automatically checkable}. Now the programmer need not
resort to manually figuring out and inserting \texttt{noalias} annotations and
worrying about what variables can and cannot be written to, re-used or freed;
instead they can let the loyal and tireless compiler do the heavy lifting.

@@

An if-expression's condition may evaluate either way during execution, so a
programmer must guarantee that both branches use the same set of linear
variables. Writing the type checker in a pure-functional monadic style paid off
here because I could now sandwich monadic values with state adjustments either
side of it. Given two monadic values that represented type checking two
branches of an if-expression, I could use the code in
Figure~\ref{fig:same_resources} to easily save, reset and compare the state
either side of running those monadic values.

@@

    \caption{Implementation of same\_resources helper method for type checking
        if-expressions. Note the monadic style helped compose computations that
        affected the type checker's state in a simple manner.}\label{fig:same_resources}

@@

A non-recursive function may be used more than once if it does not refer to any
linear variables from the surrounding scope. So, we can desugar something like
\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
let Many f =  Many (fun y : !int -> x + y) in  <body>}.  Recursion is slightly
more complicated: we can desugar the factorial function
(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int, if
(*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many}, must also
not use any linear variables from its surrounding scope, because a
recursive function may evaluate its body multiple times.

@@

In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
implementing a \emph{Kalman filter}~\cite{kalman}, a powerful set of equations
(Figure~\ref{fig:maths_kalman}) applicable to a wide variety of problems. From
the comments, we see that every variable is annotated with the matrix
expression that it will hold at some point during the computation (an
equivalent alternative, say in C++, could be to have a meaningful name for each
step and manually keep track of which names alias the same location).

@@

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed or written-over;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking values are neither used after they are freed/written-over \emph{nor} leaked.
\end{itemize}

Indeed, an inexperienced programmer could take the  na\"ive approach of just
copying sub-expressions by default and then letting the compiler tell them
which copies are never used and removing them systematically until their
program type checks.  While it is not quite a black-box, push-button
compilation of an expression, I would argue it is just as easy (if not easier)
to learn how to work with as Rust's borrow checker.

\begin{figure}[tp]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA. In contrast with the Fortran
        (Figure~\ref{fig:fortran_kalman}) or C (Figure~\ref{fig:cblas_kalman})
        implementations, this one is readable (equational and algebraic),
        using easy to reason about, value-semantic expressions. In addition to
        that, its \emph{intensional} behaviour and assumptions (regarding
        aliasing, read/write permissions and memory allocation, re-use and
        de-allocation) are also explicitly and precisely described and
        \emph{automatically checkable}.}\label{fig:ltfla_kalman}
\end{figure}

@@

All of the syntax in Figure~\ref{fig:ltfla_kalman} can be unambiguously
desugared \emph{before} type checking, through fairly simple pattern matching.
Matrix expression compilation is well-trodden territory in
academia~\cite{rocklin_thesis, fabregat_thesis, gunnels_flame, linnea, taco}
but this is, to my knowledge, the \textbf{first type-based approach} to it.

@@

One of the first hurdles I encountered was compositionality: to compile $AB +
C$ it is typically better to use a BLAS routine directly rather than first
compiling $A$, then $B$, then adding a call to multiply them, then compiling
$C$ and finishing with a call to add the results.

Another compositionality problem is that a call to a linear function does not
just return a result, but a sequence of re-bindings which dictate which
variables are still usable (in scope).  As such, to compile an expression, you
need to provide a CPS-style function of type $var \rightarrow exp$ representing
how you would use a variable representing the result in the rest of the
computation.

@@

To make the code produced readable, I added a few `optimisations', to the
compiler, to `re-sugar' some of the constructs translated where appropriate.
Almost all of them involved the erasing the \mbox{!-eliminator} which is not needed in
regular, intuitionistic OCaml. So, we can simply replace any expression of
the form:

@@

    \item \ltfla{let xy, z = <exp> in let x,y = xy in <body>} with
        \ltfla{let (x,y), z = <exp> in <body>},

@@

\begin{figure}[tp]
    \centering
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let rec f i n x0 row =
  if Prim.extract @@ Prim.eqI i n then (row, x0)
  else
    let row, x1 = Prim.get row i in
    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
in
f
    \end{minted}
    \caption{Recursive OCaml function for a summing over an array, generated (at
        \emph{compile time}) from the code in Figure~\ref{fig:ltfla_sumarray},
        passed through \texttt{ocamlformat} for presentation.}\label{fig:ocaml_sumarray}

@@

It is clear that both Figures~\ref{fig:fortran_kalman}
and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the intensional
behaviour and assumptions (with respect to read/writes permissions, memory
management and aliasing) of the program and the BLAS primitives it uses \textbf{more
accurately} than a Fortran, C or OCaml implementation could.

@@

Generating code at compile time has two advantages. First, it avoids runtime
overhead. Second, it catches interface and \textbf{relevant implementation
changes} between the generated code and the code that uses it.  The second
advantage is \emph{another benefit} of \emph{embedding} LT4LA's type system
inside of OCaml's. Relevant implementation changes are caught because LT4LA's
types express some of the intensional behaviour and assumptions of its
programs.

I suspect that this approach will be very valuable to not just \emph{users} of
libraries such as Numpy or Owl, but also their \emph{implementors}: library
functions which present a safe, value-semantic interface but use unsafe,
mutating operations on the inside could now be expressed using LT4LA and gain
\textbf{safety, value-semantics and automatic checking for their implementations}.

\clearpage%
\section{Summary}

I explained how a few core features -- linearity, the \ltfla{Many}
constructor, value-restriction, fractional-capabilities with inference,
if-expressions and recursive functions -- are enough to \emph{statically
capture and automatically check} aliasing, read/write permissions, memory
allocation, re-use and deallocation of non-trivial linear algebra programs. I
also demonstrated that simple pattern-matching and desugaring provides the
potential for a \textbf{new, \emph{type-directed}} approach to matrix expression
compilation. Lastly I have shown that it is possible to use these features with
\emph{existing} languages and frameworks.\footnote{As mentioned in the previous
chapter, if the host language supports \emph{syntax-extensions}, like PPX
for OCaml, it is possible to construct LT4LA expressions \emph{from within} the
host language.}
